---
title: "Differential expression of miRNAs on test experiment to consolidate previous results"
author: "Marco Bruttini"
date: "16 november 2023"
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Intro

Pilot experiment has only Day 10 timepoint and three conditions:

- DT 0ng (**Ctl**)
- DT 120ng (**DT**)
- S961 (**S**)

Insuline Resistance controls

All miRNA expression and sample information data are located here:  

`/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-09-18 miRNA count matrix test experiment/output`

Previous results come from here:

`/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-10-16 DE miRNAs by Timepoint`

# Analysis

Only the same miRNAs from previous low-counts filtering were kept (at least an
average of  **5 raw counts (UMI deduped)** in any condition of true experiment).

## Import data

```{r import}
setwd("/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-11-16 DE miRNAs test experiment")

library(tidyverse)

cts <- "../2023-09-18 miRNA count matrix test experiment/output/miRNA_wide.csv" %>%
  read_csv2() %>%
  column_to_rownames("name") %>%
  as.matrix

cts %>% dim
cts %>% colSums
cts %>% colSums %>% barplot

# Experimental design
coldata <- "../_data/Samples.test.csv" %>%
  read_csv2(col_types = "fffffcnn", locale = locale(decimal_mark = ",")) %>%
  mutate(
    Hemolysis = Hemolysis %>%
      factor(levels = c("No", "Little", "Moderate", "Significant", "Strong")),
    Day = Day %>%
      fct_relabel(~ paste("Day", .x %>% str_pad(2, "left"))),
    Condition = Group %>%
      fct_recode(BCA="β-cell ablation", IR="Insulin resistance") %>%
      paste(Treatment) %>%
      factor,
    rowname = Sample
  ) %>%
  group_by(Condition) %>%
  arrange(Sample) %>%
  column_to_rownames()

```

## Differential expression

```{r "de norm"}
library(DESeq2)
library(ggrepel)
library(ggupset)
library(pheatmap)
library(EnhancedVolcano)
library(BiocParallel)
library(GGally)
library(viridis)
library(ggpointdensity)
library(VennDiagram)
library(ashr)
library(RColorBrewer)

register(SnowParam(12))

pointdensity <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_abline(color = "red") +
    geom_pointdensity(...) +
    scale_color_viridis("Density") +
    theme_bw() 
}

# Low-counts filtering
keep <- "../2023-10-16 DE miRNAs by Timepoint/output/Day 10.counts.raw.csv" %>%
  read_csv2(show_col_types = FALSE) %>%
  filter(miRNA %in% rownames(cts)) %>%
  pull(miRNA)

cts <- cts[keep, ]
cts %>% dim
cts %>% colSums
cts %>% colSums %>% barplot

cts %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  write_excel_csv2(paste0("output/counts.raw.csv"))

dds <- cts %>%
  DESeqDataSetFromMatrix(
    colData = coldata %>%
      select(!Day),
    design = ~ 0 + Condition
  ) %>%
  DESeq

plotDispEsts(dds)

rld <- dds %>% rlog(blind = FALSE)

rld %>% plotPCA(intgroup = "Condition")
    
rld %>%
  assay %>%
  cor %>%
  `^`(2) %>%
  pheatmap(
    annotation_col = dds %>%
      colData %>%
      as.data.frame %>%
      select(Group:`RNA.Volume`),
    show_rownames = F,
    main = "Replicates correlation"
  )
    
rld <- rld %>%
  assay() %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata, join_by(Sample)) %>%
  group_by(Group, Treatment, miRNA)
    
ggsave(
  filename = paste0("output/scatter.mean.Group.rlog.png"),
  width = 16,
  height = 9,
  plot = rld %>%
    group_by(Group, miRNA) %>%
    summarise(Reads = Reads %>% mean, .groups = "drop") %>%
    pivot_wider(id_cols = miRNA, names_from = Group, values_from = Reads) %>%
    select(!miRNA) %>%
    ggpairs(lower = list(continuous = pointdensity))
)

ggsave(
  filename = paste0("output/scatter.mean.Condition.rlog.png"),
  width = 16,
  height = 9,
  plot = rld %>%
    group_by(Condition, miRNA) %>%
    summarise(Reads = Reads %>% mean, .groups = "drop") %>%
    pivot_wider(id_cols = miRNA, names_from = Condition, values_from = Reads) %>%
    select(!miRNA) %>%
    ggpairs(lower = list(continuous = pointdensity))
)
    
rld %>%
  summarise(Reads = Reads %>% mean, .groups = "drop") %>%
  pivot_wider(names_from = Treatment, values_from = Reads) %>%
  pivot_longer(!Group:Control, names_to = "Treatment", values_to = "Treated") %>%
  pointdensity(aes(x = Control, y = Treated)) +
  facet_grid(. ~ Treatment) +
  ggtitle("rlog normalized counts")
ggsave(
  filename = paste0("output/scatter.control_treated.rlog.png"),
  width = 16,
  height = 9
)
    
normalized <- dds %>%
  counts(normalized = TRUE) %>%
  as.data.frame %>%
  rownames_to_column("miRNA")
    
normalized %>%
  mutate(across(!miRNA, round)) %>%
  write_excel_csv2(paste0("output/counts.normalized.csv"))
    
norm.means <- normalized %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Counts") %>%
  inner_join(dds %>% colData %>% as.data.frame, join_by(Sample)) %>%
  group_by(Group, Treatment, miRNA) %>%
  summarise(
    Mean = Counts %>% mean,
    SD = Counts %>% sd,
    .groups = "drop"
  )
    
norm.means %>%
  pivot_wider(names_from = Group:Treatment, values_from = Mean:SD) %>%
  mutate(across(!miRNA, round)) %>%
  write_excel_csv2(paste0("output/counts.normalized.group.mean.csv"))
    
ggsave(
  filename = paste0(paste0("output/bar.counts.normalized.top-100.png")),
  width = 18,
  height = 32,
  plot = norm.means %>%
    slice_max(Mean, n = 100) %>%
    arrange(desc(Mean)) %>%
    mutate(
      miRNA = miRNA %>% fct_inorder
    ) %>%
    ggplot(aes(x = miRNA, y = Mean)) +
    geom_col() +
    geom_errorbar(aes(ymin = pmax(0, Mean - SD), ymax = Mean + SD)) +
    facet_grid(Group + Treatment ~ .) +
    labs(title = "Top 100 expressed miRNAs by group", y = "Mean of normalized counts") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
)
    
ggsave(
  filename = paste0(paste0("output/line.counts.normalized.group.mean.png")),
  width = 18,
  height = 32,
  plot = norm.means %>%
    ggplot(aes(x = Treatment, y = log2(Mean + 1))) +
    geom_line(aes(group = miRNA), alpha = .1) +
    facet_grid(. ~ Group, scales = "free_x")
)

```

### Treatments vs Control

```{r}

de <- tribble(
    ~group, ~contrast1,     ~contrast2,    ~fileName, ~plotTitle,
    "BCA",  "BCA.DT.120ng", "BCA.Control", "BCA.DT120-vs-CTR", "BCA DT (120ng) vs Control",
    "IR",   "IR.S961",      "BCA.Control", "IR.S961-vs-BCA.CTR",   "IR S961 vs BCA Control",
  ) %>%
  rowwise %>%
  group_map(~ {
    print(.x$plotTitle)
    dds %>%
      results(
        contrast = c("Condition", .x$contrast1, .x$contrast2),
        alpha = 0.05
      ) %>%
      lfcShrink(dds = dds, res = ., type = "ashr") %>%
      {
        plotMA(.)
        summary(.)
        .
      } %>%
      as.data.frame %>%
      rownames_to_column("miRNA") %>%
      arrange(padj, pvalue) %>%
      write_excel_csv2(paste0("output/DE/DE.", .x$fileName, ".csv")) %>%
      mutate(
        group = .x$group,
        contrast1 = .x$contrast1,
        contrast2 = .x$contrast2,
        test = .x$fileName %>% str_remove("^[^.]*\\.") %>% str_replace("\\.", "_"),
        direction = if_else(log2FoldChange > 0, "UP", "DOWN"),
        .before = miRNA
      ) %>%
      {
        EnhancedVolcano(
          toptable = .,
          lab = .$miRNA,
          x = 'log2FoldChange',
          y = 'padj',
          pCutoff = 0.05,
          FCcutoff = 0,
          xlim = c(-2, 2.5),
          title = .x$plotTitle,
          subtitle = bquote(italic("BH adjusted p-values")),
          caption = paste(
            length(which(.$padj < 0.05)),
            "differentially expressed features out of",
            nrow(.),
            "total features"
          )
        )
        ggsave(
          paste0("output/DE/DE.", .x$fileName, ".volcano.png"),
          width = 16,
          height = 9
        )
        .
      }
  }) %>%
  bind_rows

de %>%
  filter(padj < 0.05) %>%
  group_by(group, contrast1, contrast2, test) %>%
  group_walk(~ {
    ggsave(
      paste0("output/DE/DE.", .y$group, ".", .y$test, ".boxplot.png"),
      width = 16,
      height = 9,
      plot = .x %>%
        select(miRNA, direction) %>%
        inner_join(
          dds %>%
            counts(normalized = T) %>%
            as.data.frame %>%
            rownames_to_column("miRNA"),
          join_by(miRNA)
        ) %>%
        pivot_longer(!miRNA:direction, names_to = "Sample", values_to = "Reads") %>%
        inner_join(coldata, join_by(Sample)) %>%
        filter(
          Condition %>% make.names %in% c(.y$contrast1, .y$contrast2)
        ) %>%
        ggplot(aes(x = Condition, y = log2(Reads + 1))) +
        stat_summary(aes(group = miRNA), fun = mean, geom = "line", linetype = "dashed", alpha = .5) +
        geom_boxplot(aes(fill = Condition)) + #aes(fill = Day), position = position_dodge2(padding = 0.2)) +
        facet_wrap(~ direction + miRNA, scales = "free_y") +
        scale_fill_manual(values = brewer.pal(6, "RdYlBu")[c(5, 1)]) +
        #scale_fill_manual(values = brewer.pal(6, "RdYlBu")[3:1]) +
        scale_x_discrete(expand = expansion(add = 0.4)) +
        theme_bw() +
        theme(legend.position = "none")
    )
  })

prev.de <- de %>%
  mutate(exp = "Pilot") %>%
  select(exp, group, miRNA:padj)

prev.de <- "../2023-10-16 DE miRNAs by Timepoint/output/DE/DE.BCA.DT120-vs-CTR.Day10.csv" %>%
  read_csv2(show_col_types = FALSE) %>%
  mutate(exp = "Big", group = "BCA", .before = miRNA) %>%
  bind_rows(prev.de)

prev.de <- "../2023-10-16 DE miRNAs by Timepoint/output/DE/DE.IR.S961-vs-CTR.Day10.csv" %>%
  read_csv2(show_col_types = FALSE) %>%
  mutate(exp = "Big", group = "IR", .before = miRNA) %>%
  bind_rows(prev.de)

prev.de <- prev.de %>%
  pivot_wider(id_cols = group:miRNA, names_from = exp, values_from = baseMean:padj) %>%
  write_csv2("output/DE/DE.comparison.csv")

prev.de %>%
  filter(log2FoldChange_Pilot < 20) %>%
  group_by(group) %>%
  group_walk(~ {
    .x %>%
      ggplot(aes(x = log2FoldChange_Pilot, y = log2FoldChange_Big)) +
      geom_hline(yintercept = 0, color = "grey") +
      geom_vline(xintercept = 0, color = "grey") +
      geom_pointdensity(aes(size = baseMean_Big)) +
      geom_point(data = .x %>% filter(padj_Big < 0.05), color = "grey", shape = 1, size = 8) +
      geom_text_repel(data = .x %>% filter(padj_Big < 0.05), aes(label = miRNA), box.padding = 1, seed = 123) +
      geom_point(data = .x %>% filter(padj_Pilot < 0.05), color = "red", shape = 3, size = 7) +
      scale_color_viridis("Density") +
      scale_size("Mean RC (Big)", trans = "log10", labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
      theme_bw() +
      xlab("log2FoldChange (Pilot)") +
      ylab("log2FoldChange (Big)") +
      guides(
        size = guide_legend(order = 1),
        color = guide_colorbar(order = 2)
      )
    ggsave(paste0("output/DE/DE.comparison.", .y$group, ".png"), width = 12, height = 9)
  })

```


