---
title: "Differential expression of miRNAs within the same collection Day"
author: "Marco Bruttini"
date: "16 october 2023"
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Intro

All miRNA expression and sample information data are located here:  

`/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-09-05 miRNA count matrix/output`

Longitudinal analysis may be impaired by a strong batch processing effect. See
_2023-09-15 miRNA profiling.pptx_ report.

# Analysis

A minimal low counts filtering was applied to both groups:

- miRNAs with at least an average of  **5 raw counts (UMI deduped)** in any
condition were kept.

## Import data

```{r import}
setwd("/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-10-16 DE miRNAs by Timepoint")

library(tidyverse)

cts <- "../2023-09-05 miRNA count matrix/output/miRNA_wide.csv" %>%
  read_csv2() %>%
  column_to_rownames("name") %>%
  as.matrix

# Experimental design
coldata <- "../_data/Samples.experiment.csv" %>%
  read_csv2(col_types = "fffffcn", locale = locale(decimal_mark = ",")) %>%
  mutate(
    Hemolysis = Hemolysis %>%
      factor(levels = c("No", "Little", "Moderate", "Significant", "Strong")),
    Day = Day %>%
      fct_relabel(~ paste("Day", .x %>% str_pad(2, "left"))),
    Condition = Group %>%
      fct_recode(BCA="β-cell ablation", IR="Insulin resistance") %>%
      paste(Treatment) %>%
      factor,
    rowname = Sample
  ) %>%
  group_by(Condition) %>%
  arrange(Sample) %>%
  column_to_rownames()

```

## Differential expression

```{r "de norm"}
library(DESeq2)
library(ggrepel)
library(ggupset)
library(pheatmap)
library(EnhancedVolcano)
library(BiocParallel)
library(GGally)
library(viridis)
library(ggpointdensity)
library(VennDiagram)
library(ashr)
library(RColorBrewer)

register(SnowParam(12))

pointdensity <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_abline(color = "red") +
    geom_pointdensity(...) +
    scale_color_viridis("Density") +
    theme_bw() 
}

# Low-counts filtering
keep <- cts %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata, join_by(Sample)) %>%
  group_by(Group, Treatment, Day, miRNA) %>%
  summarise(Reads = Reads %>% mean) %>%
  group_by(miRNA) %>%
  summarise(Reads = Reads %>% max) %>%
  filter(Reads >= 10) %>%
  pull(miRNA)

dds <- coldata %>%
  pull(Day) %>%
  levels %>%
  set_names %>%
  map(function(timepoint) {
  
    print(timepoint)
    
    # Splitting by timepoint
    tp_cts <- cts[keep, coldata$Day == timepoint]
    tp_cts %>% dim
    tp_cts %>% colSums
    tp_cts %>% colSums %>% barplot
    
    tp_cts %>%
      as.data.frame %>%
      rownames_to_column("miRNA") %>%
      write_excel_csv2(paste0("output/", timepoint, ".counts.raw.csv"))

    DESeqDataSetFromMatrix(
      countData = tp_cts,
      colData = coldata %>%
        filter(Day == timepoint) %>%
        select(!Day),
      design = ~ 0 + Condition
    ) %>%
      DESeq
  })

dds %>%
  iwalk(function(dds, timepoint) {
  
    print(timepoint)
    
    plotDispEsts(dds)

    rld <- dds %>% rlog(blind = FALSE)

    rld %>% plotPCA(intgroup = "Condition")
    
    rld %>%
      assay %>%
      cor %>%
      `^`(2) %>%
      pheatmap(
        annotation_col = dds %>%
          colData %>%
          as.data.frame %>%
          select(Group:`RNA.Concentration`),
        show_rownames = F,
        main = "Replicates correlation"
      )
    
    rld <- rld %>%
      assay() %>%
      as.data.frame %>%
      rownames_to_column("miRNA") %>%
      pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
      inner_join(coldata, join_by(Sample)) %>%
      group_by(Group, Treatment, miRNA)
    
    ggsave(
      filename = paste0("output/", timepoint, ".scatter.mean.Group.rlog.png"),
      width = 16,
      height = 9,
      plot = rld %>%
        group_by(Group, miRNA) %>%
        summarise(Reads = Reads %>% mean, .groups = "drop") %>%
        pivot_wider(id_cols = miRNA, names_from = Group, values_from = Reads) %>%
        select(!miRNA) %>%
        ggpairs(lower = list(continuous = pointdensity))
    )

    ggsave(
      filename = paste0("output/", timepoint, ".scatter.mean.Condition.rlog.png"),
      width = 16,
      height = 9,
      plot = rld %>%
        group_by(Condition, miRNA) %>%
        summarise(Reads = Reads %>% mean, .groups = "drop") %>%
        pivot_wider(id_cols = miRNA, names_from = Condition, values_from = Reads) %>%
        select(!miRNA) %>%
        ggpairs(lower = list(continuous = pointdensity))
    )
    
    rld %>%
      summarise(Reads = Reads %>% mean, .groups = "drop") %>%
      pivot_wider(names_from = Treatment, values_from = Reads) %>%
      pivot_longer(!Group:Control, names_to = "Treatment", values_to = "Treated") %>%
      pointdensity(aes(x = Control, y = Treated)) +
      facet_grid(. ~ Treatment) +
      ggtitle("rlog normalized counts")
    ggsave(
      filename = paste0("output/", timepoint, ".scatter.control_treated.rlog.png"),
      width = 16,
      height = 9
    )
    
    normalized <- dds %>%
      counts(normalized = TRUE) %>%
      as.data.frame %>%
      rownames_to_column("miRNA")
    
    normalized %>%
      mutate(across(!miRNA, round)) %>%
      write_excel_csv2(paste0("output/", timepoint, ".counts.normalized.csv"))
    
    norm.means <- normalized %>%
      pivot_longer(!miRNA, names_to = "Sample", values_to = "Counts") %>%
      inner_join(dds %>% colData %>% as.data.frame, join_by(Sample)) %>%
      group_by(Group, Treatment, miRNA) %>%
      summarise(
        Mean = Counts %>% mean,
        SD = Counts %>% sd,
        .groups = "drop"
      )
    
    norm.means %>%
      pivot_wider(names_from = Group:Treatment, values_from = Mean:SD) %>%
      mutate(across(!miRNA, round)) %>%
      write_excel_csv2(paste0("output/", timepoint, ".counts.normalized.group.mean.csv"))
    
    ggsave(
      filename = paste0(paste0("output/", timepoint, ".bar.counts.normalized.top-100.png")),
      width = 18,
      height = 32,
      plot = norm.means %>%
        slice_max(Mean, n = 100) %>%
        arrange(desc(Mean)) %>%
        mutate(
          miRNA = miRNA %>% fct_inorder
        ) %>%
        ggplot(aes(x = miRNA, y = Mean)) +
        geom_col() +
        geom_errorbar(aes(ymin = pmax(0, Mean - SD), ymax = Mean + SD)) +
        facet_grid(Group + Treatment ~ .) +
        labs(title = "Top 100 expressed miRNAs by group", y = "Mean of normalized counts") +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
    )
    
    ggsave(
      filename = paste0(paste0("output/", timepoint, ".line.counts.normalized.group.mean.png")),
      width = 18,
      height = 32,
      plot = norm.means %>%
        ggplot(aes(x = Treatment, y = log2(Mean + 1))) +
        geom_line(aes(group = miRNA), alpha = .1) +
        facet_grid(. ~ Group, scales = "free_x")
    )
    
  })

```

### Treatments vs Control at each Timepoint

```{r}

de <- dds %>%
  imap(function(dds, timepoint) {
    
    tribble(
      ~group, ~contrast1,     ~contrast2,    ~fileName, ~plotTitle,
      "BCA",  "BCA.DT.005ng", "BCA.Control", "BCA.DT005-vs-CTR", "BCA DT (5ng) vs Control",
      "BCA",  "BCA.DT.015ng", "BCA.Control", "BCA.DT015-vs-CTR", "BCA DT (15ng) vs Control",
      "BCA",  "BCA.DT.120ng", "BCA.Control", "BCA.DT120-vs-CTR", "BCA DT (120ng) vs Control",
      "IR",   "IR.S961",      "IR.Control",  "IR.S961-vs-CTR",   "IR S961 vs Control",
    ) %>%
      mutate(
        day = timepoint,
        fileName = fileName %>%
          paste0(".", timepoint %>% str_replace_all(c("  " = "0", " " = ""))),
        plotTitle = plotTitle %>%
          paste("at", timepoint %>% str_replace("  ", " "))
      ) %>%
      rowwise %>%
      group_map(~ {
        print(.x$plotTitle)
        dds %>%
          results(
            contrast = c("Condition", .x$contrast1, .x$contrast2),
            alpha = 0.05
          ) %>%
          lfcShrink(dds = dds, res = ., type = "ashr") %>%
          {
            plotMA(.)
            summary(.)
            .
          } %>%
          as.data.frame %>%
          rownames_to_column("miRNA") %>%
          arrange(padj, pvalue) %>%
          write_excel_csv2(paste0("output/DE/DE.", .x$fileName, ".csv")) %>%
          mutate(
            group = .x$group,
            day = .x$day,
            contrast1 = .x$contrast1,
            contrast2 = .x$contrast2,
            test = .x$fileName %>% str_remove("^[^.]*\\."),
            direction = if_else(log2FoldChange > 0, "UP", "DOWN"),
            .before = miRNA
          ) %>%
          {
            EnhancedVolcano(
              toptable = .,
              lab = .$miRNA,
              x = 'log2FoldChange',
              y = 'padj',
              pCutoff = 0.05,
              FCcutoff = 0,
              title = .x$plotTitle,
              subtitle = bquote(italic("BH adjusted p-values")),
              caption = paste(
                length(which(.$padj < 0.05)),
                "differentially expressed features out of",
                nrow(.),
                "total features"
              )
            )
            ggsave(
              paste0("output/DE/DE.", .x$fileName, ".volcano.png"),
              width = 16,
              height = 9
            )
            .
          }
      }) %>%
      bind_rows
  }) %>%
  bind_rows

de %>%
  filter(padj < 0.05) %>%
  group_by(group, day, contrast1, contrast2, test) %>%
  group_walk(~ {
    ggsave(
      paste0("output/DE/DE.", .y$group, ".", .y$test, ".boxplot.png"),
      width = 16,
      height = 9,
      plot = .x %>%
        select(miRNA, direction) %>%
        inner_join(
          dds[[.y$day]] %>%
            counts(normalized = T) %>%
            as.data.frame %>%
            rownames_to_column("miRNA"),
          join_by(miRNA)
        ) %>%
        pivot_longer(!miRNA:direction, names_to = "Sample", values_to = "Reads") %>%
        inner_join(coldata, join_by(Sample)) %>%
        filter(
          Condition %>% make.names %in% c(.y$contrast1, .y$contrast2)
        ) %>%
        ggplot(aes(x = Condition, y = log2(Reads + 1))) +
        stat_summary(aes(group = miRNA), fun = mean, geom = "line", linetype = "dashed", alpha = .5) +
        geom_boxplot(aes(fill = Condition)) + #aes(fill = Day), position = position_dodge2(padding = 0.2)) +
        facet_wrap(~ direction + miRNA, scales = "free_y") +
        ggtitle(.y$day) +
        scale_fill_manual(values = brewer.pal(6, "RdYlBu")[c(5, 1)]) +
        #scale_fill_manual(values = brewer.pal(6, "RdYlBu")[3:1]) +
        scale_x_discrete(expand = expansion(add = 0.4)) +
        theme_bw() +
        theme(legend.position = "none")
    )
  })

de %>%
  filter(padj < 0.05) %>%
  mutate(test = paste0(test, ".", direction)) %>%
  group_by(group, miRNA) %>%
  summarise(tests = paste(test, collapse = "; ")) %>%
  group_walk(~ {
    .x %>%
    write_excel_csv2(paste0("output/DE/DE.intersections.", .y$group, ".FDR.csv"))
  })

de %>%
  filter(padj < 0.05) %>%
  mutate(test = paste0(test, ".", direction)) %>%
  group_by(group) %>%
  group_walk(~ {
    .x %>%
      group_by(miRNA) %>%
      summarise(tests = list(test)) %>%
      {
        ggplot(., aes(x = tests)) +
          geom_bar() +
          geom_text(stat = "count", aes(label = after_stat(count), y = after_stat(count) / 2), color = "white") +
          scale_x_upset(name = "Differential expression", sets = .$tests %>% unlist %>% sort %>% unique %>% rev) +
          scale_y_continuous("Count", labels = scales::label_number(accuracy = 1)) +
          ggtitle("All miRNAs with FDR < 0.05", .y$group)
      } %>%
      ggsave(
        filename = paste0("output/DE/DE.intersections.", .y$group, ".FDR.png"),
        plot = .,
        width = 16,
        height = 9
      )
  })

de %>%
  filter(pvalue < 0.05) %>%
  mutate(test = paste0(test, ".", direction)) %>%
  group_by(group, miRNA) %>%
  summarise(tests = paste(test, collapse = "; ")) %>%
  group_walk(~ {
    .x %>%
    write_excel_csv2(paste0("output/DE/DE.intersections.", .y$group, ".pvalue.csv"))
  })

de %>%
  filter(pvalue < 0.05) %>%
  mutate(test = factor(paste0(test, ".", direction))) %>%
  group_by(group) %>%
  group_walk(~ {
    .x %>%
      group_by(miRNA) %>%
      summarise(tests = list(test)) %>%
      {
        ggplot(., aes(x = tests)) +
          geom_bar() +
          geom_text(stat = "count", aes(label = after_stat(count)), vjust = -1, size = 3) +
          scale_x_upset(name = "Differential expression", sets = .$tests %>% unlist %>% sort %>% unique %>% rev) +
          scale_y_continuous("Count", labels = scales::label_number(accuracy = 1), limits = c(0, 40)) +
          ggtitle("All miRNAs with p value < 0.05", .y$group)
      } %>%
      ggsave(
        filename = paste0("output/DE/DE.intersections.", .y$group, ".pvalue.png"),
        plot = .,
        width = 16,
        height = 9
      )
  })

de %>%
  filter(pvalue < 0.05) %>%
  mutate(test = factor(paste0(test, ".", direction))) %>%
  group_by(group) %>%
  group_walk(~ {
    .x %>%
      group_by(miRNA) %>%
      summarise(tests = list(test)) %>%
      {
        ggplot(., aes(x = tests)) +
          geom_bar() +
          geom_text(stat = "count", aes(label = after_stat(count)), vjust = -1, size = 3) +
          scale_x_mergelist(sep = "|", limits = rev) +
          axis_combmatrix(sep = "\\|", override_plotting_function = function(df) {
            df %>%
              ggplot(aes(x = at, y = single_label)) +
              geom_rect(aes(fill = index %% 2 == 0), ymin = df$index - 0.5, ymax = df$index + 0.5, xmin = 0, xmax = 1) +
              geom_line(data = function(dat) dat[dat$observed, , drop = FALSE], aes(group = labels), linewidth = 1) +
              geom_point(data = function(dat) dat[dat$observed, , drop = FALSE], 
                aes(
                  color = single_label %>% str_extract("^[^-]*"),
                  alpha = single_label %>% str_extract("\\.(Day\\d\\d)\\.", group = 1),
                  shape = single_label %>% str_extract("(UP|DOWN)$")
                ),
                size = 3
              ) +
              ylab("") + xlab("") +
              scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
              scale_y_discrete(limits = rev) +
              scale_shape_manual(values = c(UP = 24, DOWN = 25)) +
              scale_fill_manual(values = c(`TRUE` = "white", `FALSE` = "#F7F7F7")) +
              scale_alpha_manual(values = c(Day00 = .33, Day04 = .67, Day10 = 1)) +
              guides(fill = "none", color = "none", alpha = "none", shape = "none") +
              theme(
                legend.position = "bottom",
                panel.background = element_blank(),
                axis.text.x = element_blank(),
                axis.ticks.y = element_blank(),
                axis.ticks.length = unit(0, "pt"),
                axis.title.y = element_blank(),
                axis.title.x = element_blank(),
                axis.line = element_blank(),
                panel.border = element_blank()
              )
          }) +
          scale_y_continuous("Count", labels = scales::label_number(accuracy = 1), limits = c(0, 40)) +
          ggtitle("All miRNAs with p value < 0.05", .y$group)
      } %>%
      ggsave(
        filename = paste0("output/DE/DE.intersections.", .y$group, ".pvalue.colors.png"),
        plot = .,
        width = 16,
        height = 9
      )
  })

de %>%
  mutate(test = factor(paste0(test, ".", direction))) %>%
  group_by(group) %>%
  group_walk(~ {
    .x %>%
      group_by(miRNA) %>%
      summarise(tests = list(test)) %>%
      {
        ggplot(., aes(x = tests)) +
          geom_bar() +
          geom_text(stat = "count", aes(label = after_stat(count)), vjust = -1, size = 3) +
          scale_x_upset(name = "Differential expression", sets = .$tests %>% unlist %>% sort %>% unique %>% rev) +
          scale_y_continuous("Count", labels = scales::label_number(accuracy = 1)) +
          ggtitle("All miRNAs", .y$group)
      } %>%
      ggsave(
        filename = paste0("output/DE/DE.intersections.", .y$group, ".no-filter.png"),
        plot = .,
        width = 32,
        height = 18
      )
  })

de %>%
  group_by(miRNA) %>%
  filter(max(abs(log2FoldChange) - lfcSE) > 0) %>%
  group_walk(~ {
    ggsave(
      filename = paste0("output/DE/miRNAs/", .y$miRNA, ".png"),
      plot = .x %>%
        ggplot(aes(x = day, y = log2FoldChange, color = contrast1)) +
        geom_hline(yintercept = 0) +
        geom_line(aes(group = contrast1), position = position_dodge2(width = .5), linewidth = 1) +
        geom_pointrange(
          aes(
            ymin = log2FoldChange - lfcSE,
            ymax = log2FoldChange + lfcSE,
            shape = abs(log2FoldChange) - lfcSE > 0
          ),
          fill = "white",
          position = position_dodge2(width = .5)
        ) +
        geom_text(aes(label = if_else(padj < 0.05, "*", "")), hjust = -1, vjust = .75, color = "black") +
        scale_color_manual(values = brewer.pal(9, "RdBu")[c(4, 3, 1, 9)]) +
        scale_shape_manual(values = c(21, 19)) +
        scale_y_continuous(limits = c(-1, 1) * max(1, abs(.x$log2FoldChange) + .$lfcSE)) +
        facet_grid(~ contrast1) +
        ggtitle(.y$miRNA, "Difference in expression respect to control group. (*: FDR < 0.05)") +
        xlab(NULL) +
        theme_bw() +
        theme(legend.position = "none"),
      width = 8,
      height = 4.5
    )
  })

```


