---
title: "Profile sRNAbench data"
author: "Marco Bruttini"
date: "5 september 2023"
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Intro

Gather together all sRNAbench data from samples zip folders.

Although the columns in output files report the name "readCount", they're
actually UMI-corrected as sRNAbench performs UMI collapsing at preprocessing
step when QIASeq library is selected.

# Analysis

A minimal low counts filtering is applied to all sncRNA species.
Features with at least **5 raw counts (UMI deduped)** in any sample are retained.

## Import data

```{r import}
setwd("/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-09-05 miRNA count matrix")

library(tidyverse)

fix_composite_names <- function(names_to_fix) {
  names_to_fix %>%
    modify(
      ~ str_split(., "\\$") %>%
        unlist %>%
        str_sort(numeric = TRUE) %>%
        paste(collapse = "$")
    )
}

miRNA <- other <- results <- tibble()
minRC <- 5

dir.create(path = "input/temp")

  for (s in list.files(path = "../_data/sRNAbench/Experiment", pattern = "*.zip")) {
    
    s <- s %>% str_remove("\\.zip$")
    
    message(s)
    
    unzip(
      zipfile = paste0("../_data/sRNAbench/Experiment/", s, ".zip"),
      files = file.path(s, c(
        "results.txt",
        "stat/mappingStat.txt",
        "mature_sense.grouped",
        "GRCm39_genomic_tRNA_classificationFull.txt",
        "GRCm39_ncRNA_sense.grouped",
        "GRCm39_RNAcentral_sense.grouped"
      )),
      junkpaths = T,
      overwrite = T,
      exdir = "input/temp"
    )
    
    results <- "input/temp/results.txt" %>%
      read_lines() %>%
      enframe(name = NULL) %>%
      filter(
        str_detect(value, "^reads(=|Raw|AdapterNotFound|LengthFilteredM(in|ax)|(RCcomplexityF|(Q?RC|UMI)f)iltered)|^shortUMI|^(total|(unA|a)ssigned)RC")
      ) %>%
      separate(value, c("variable", "value"), "=", convert = T) %>%
      mutate(sample = s, .before = 1) %>%
      bind_rows(results)
    
    results <- "input/temp/mappingStat.txt" %>%
      read_tsv(col_types = "c--d---") %>%
      filter(
        name %in% paste0(c("mature", "tRNA", "rRNA", "lncRNA", "protein_coding", "piRNA"), "#sense")
      ) %>%
      mutate(
        variable = name %>%
          str_remove("#sense") %>%
          str_replace("mature", "miRNA") %>%
          str_replace("protein_coding", "mRNA")
      ) %>%
      group_by(variable) %>%
      summarise(
        value = RC %>% sum %>% round %>% as.integer
      ) %>%
      mutate(
        sample = s,
        .before = 1
      ) %>%
      bind_rows(results)
    
    #miRNA
    miRNA <- "input/temp/mature_sense.grouped" %>%
      read_tsv(col_types = "c--n-----") %>%
      rename(readCount = `read count (mult. map. adj.)`) %>%
      mutate(name = name %>% fix_composite_names) %>%
      group_by(name) %>%
      summarise(
        sample = s,
        readCount = readCount %>% sum %>% round %>% as.integer
      ) %>%
      bind_rows(miRNA) %>%
      relocate(name, .before = "readCount")
    
    #other (1)
    other <- "input/temp/GRCm39_genomic_tRNA_classificationFull.txt" %>%
      read_tsv(col_types = "c--n-") %>%
      rename(readCount = `RCadj`) %>%
      mutate(
        sample = s,
        class = "tRNA",
        origin = "tRNA",
        .before = 1
      ) %>%
      bind_rows(other)
    
    #other (2)
    other <- "input/temp/GRCm39_ncRNA_sense.grouped" %>%
      read_tsv(col_types = "c--n-----") %>%
      rename(readCount = `read count (mult. map. adj.)`) %>%
      mutate(
        sample = s,
        class = name %>% str_remove(".*:"),
        origin = "ncRNA",
        name = name %>% str_remove(":.*"),
        .before = 1
      ) %>%
      bind_rows(other)
    
    #other (3)
    other <- "input/temp/GRCm39_RNAcentral_sense.grouped" %>%
      read_tsv(col_types = "c--n-----") %>%
      rename(readCount = `read count (mult. map. adj.)`) %>%
      mutate(
        sample = s,
        class = name %>% str_remove_all("^[^:]*:|:[^:]*$"),
        origin = "RNAcentral",
        name = name %>% str_remove(":.*"),
        .before = 1
      ) %>%
      bind_rows(other)
    
    #other (final)
    other <- other %>%
      mutate(
        class = class %>%
          str_extract("^(lnc|pi|t|r|m)RNA$") %>%
          str_replace_na("otherRNA"),
        readCount = readCount %>% round %>% as.integer
      )
    
  }

unlink(x = "input/temp", recursive = TRUE)
rm(s)

# Experimental design
coldata <- "../_data/Samples.experiment.csv" %>%
  read_csv2(col_types = "fffffcn", locale = locale(decimal_mark = ",")) %>%
  mutate(
    Hemolysis = Hemolysis %>%
      factor(levels = c("No", "Little", "Moderate", "Significant", "Strong")),
    Day = Day %>%
      fct_relabel(~ paste("Day", .x %>% str_pad(2, "left"))),
    rowname = Sample
  ) %>%
  column_to_rownames()

```

# Results

```{r results}
results <- results %>%
  pivot_wider(names_from = "variable", values_fill = 0) %>%
  mutate(
    sample = sample %>% fct_inseq,
    unmapped = reads - totalRC,
    otherRNA = assignedRC - rowSums(across(ends_with("RNA")))
  ) %>%
  pivot_longer(!sample, names_to = "variable") %>%
  group_by(sample, variable) %>%
  summarise(value = sum(value)) %>%
  write_excel_csv2("output/results_long.csv")

results %>%
  pivot_wider(names_from = "sample", values_from = "value") %>%
  write_excel_csv2("output/results_wide.csv")

results <- results %>%
  inner_join(coldata, by = c(sample = "Sample"))

results %>%
  filter(variable == "readsRaw") %>%
  ggplot() +
  geom_col(aes(x = sample, y = value)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position = "bottom"
  ) +
  xlab("Samples") +
  ylab("Reads count")

results %>%
  filter(
    variable %>% str_detect("^reads(Raw|$)|(total|assigned)RC", negate = TRUE)
  ) %>%
  ggplot() +
  geom_col(aes(x = sample, y = value, fill = variable)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position = "bottom"
  ) +
  xlab("Samples") +
  ylab("Reads count")

results %>%
  filter(
    variable %>% str_detect("^reads(Raw|$)|RNA|totalRC", negate = TRUE)
  ) %>%
  mutate(
    variable = variable %>%
      fct_rev %>%
      fct_relevel("assignedRC", "unAssignedRC")
  ) %>%
  ggplot() +
  geom_col(aes(x = sample, y = value, fill = variable)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  scale_fill_brewer("Feature", palette = "Paired") +
  xlab("Samples") +
  ylab("Reads count")
#ggsave("output/results.qc.pdf", width = 16, height = 9)
ggsave("output/results.qc.png", width = 16, height = 9)

results %>%
  filter(
    variable %>% str_detect("^reads(Raw|$)|RNA|totalRC", negate = TRUE)
  ) %>%
  mutate(
    variable = variable %>%
      fct_rev %>%
      fct_relevel("assignedRC", "unAssignedRC")
  ) %>%
  ggplot() +
  geom_col(aes(x = sample, y = value, fill = variable), position = position_fill()) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  scale_fill_brewer("Feature", palette = "Paired") +
  scale_y_continuous(labels = scales::percent) +
  xlab("Samples") +
  ylab("Reads count")
#ggsave("output/results.qc.ratio.pdf", width = 16, height = 9)
ggsave("output/results.qc.ratio.png", width = 16, height = 9)

results %>%
  filter(
    variable %>% str_detect("^reads(Raw|$)|RNA|totalRC", negate = TRUE)
  ) %>%
  group_by(sample) %>%
  mutate(value = value / sum(value)) %>%
  group_by(variable) %>%
  summarise(ratio = value %>% mean) %>%
  write_excel_csv2("output/results.qc.ratio.csv")

results %>%
  filter(
    variable %>% str_detect("RNA")
  ) %>%
  ggplot() +
  geom_col(aes(x = sample, y = value, fill = variable)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  scale_fill_brewer("Feature", palette = "Paired") +
  xlab("Samples") +
  ylab("Reads count")
#ggsave("output/results.RNA.pdf", width = 16, height = 9)
ggsave("output/results.RNA.png", width = 16, height = 9)

results %>%
  filter(
    variable %>% str_detect("RNA")
  ) %>%
  ggplot() +
  geom_col(aes(x = sample, y = value, fill = variable), position = position_fill()) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  scale_fill_brewer("Feature", palette = "Paired") +
  scale_y_continuous(labels = scales::percent) +
  xlab("Samples") +
  ylab("Reads count")
#ggsave("output/results.RNA.ratio.pdf", width = 16, height = 9)
ggsave("output/results.RNA.ratio.png", width = 16, height = 9)

# results %>%
#   filter(
#     variable %>% str_detect("RNA")
#   ) %>%
#   group_by(sample) %>%
#   mutate(value = value / sum(value)) %>%
#   write_excel_csv2("output/results.RNA.ratio.noAvg.csv") %>%
#   group_by(variable) %>%
#   summarise(ratio = value %>% mean) %>%
#   write_excel_csv2("output/results.RNA.ratio.csv")

```

## miRNA

```{r miRNA}
miRNA <- miRNA %>%
  mutate(sample = sample %>% fct_inseq) %>%
  arrange(sample, name) %>%
  pivot_wider(names_from = "sample", values_from = "readCount", values_fill = 0) %>%
  rowwise %>%
  filter(max(c_across(!name)) >= minRC) %>%
  write_excel_csv2("output/miRNA_wide.csv") %>%
  pivot_longer(!name, names_to = "sample", values_to = "readCount") %>%
  write_excel_csv2("output/miRNA_long.csv")

miRNA %>%
  filter(readCount >= minRC) %>%
  count(sample) %>%
  write_excel_csv2("output/miRNA.count.csv")%>%
  inner_join(coldata, by = c(sample = "Sample")) %>%
  ggplot() +
  geom_col(aes(x = sample, y = n)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  xlab("Samples") +
  ylab("miRNA count")
#ggsave("output/miRNA.count.pdf", width = 16, height = 9)
ggsave("output/miRNA.count.png", width = 16, height = 9)

miRNA %>%
  filter(readCount >= minRC) %>%
  count(sample) %>%
  inner_join(coldata, by = c(sample = "Sample")) %>%
  group_by(Day) %>%
  summarise(n = n %>% mean %>% round %>% as.integer) %>%
  write_excel_csv2("output/miRNA.count.average.csv")

miRNA %>%
  ggplot() +
  geom_density(aes(x = log2(readCount + 1), group = sample))
#ggsave("output/miRNA.distribution.pdf", width = 16, height = 9)
#ggsave("output/miRNA.distribution.png", width = 16, height = 9)

miRNA %>%
  inner_join(coldata, by = c(sample = "Sample")) %>%
  group_by(Group, Treatment, Day, name) %>%
  summarise(
    meanRC = readCount %>% mean %>% round %>% as.integer,
    sdRC = readCount %>% sd %>% round %>% as.integer
  ) %>%
  slice_max(meanRC, n = 10, with_ties = FALSE) %>%
  mutate(
    name = name %>% str_remove("^mmu-"),
    rank = 1:10 %>% factor,
    .after = name
  ) %>%
  write_excel_csv2("output/miRNA.top10.csv") %>%
  ggplot() +
  geom_col(aes(x = rank, y = log2(meanRC + 1), fill = name)) +
  geom_label(aes(x = rank, y = 0, label = name), vjust = 0) +
  facet_grid(rows = vars(Group, Treatment, Day), switch = "y") +
  scale_fill_viridis_d() +
  theme_bw() +
  guides(fill = "none") +
  xlab("Rank")
#ggsave("output/miRNA.top10.pdf", width = 18, height = 32)
ggsave("output/miRNA.top10.png", width = 18, height = 32)

```

## Other

```{r other}

other <- other %>%
  mutate(sample = sample %>% fct_inseq) %>%
  arrange(sample, class, origin, name) %>%
  pivot_wider(names_from = "sample", values_from = "readCount", values_fill = 0) %>%
  arrange(class, origin, name) %>%
  rowwise %>%
  filter(max(c_across(!class:name)) >= minRC) %>%
  write_excel_csv2("output/other_wide.csv") %>%
  pivot_longer(!class:name, names_to = "sample", values_to = "readCount") %>%
  write_excel_csv2("output/other_long.csv")

other %>%
  filter(class == "tRNA", origin == "tRNA", readCount >= minRC) %>%
  count(sample) %>%
  write_excel_csv2("output/tRNA.count.csv")

other %>%
  filter(readCount >= minRC) %>%
  count(sample, class) %>%
  write_excel_csv2("output/other.count.csv") %>%
  inner_join(coldata, by = c(sample = "Sample")) %>%
  ggplot() +
  geom_col(aes(x = sample, y = n, fill = class)) +
  facet_wrap(~ paste(Group, Treatment, Day, sep = " - "), scales = "free_x", ncol = 6, dir = "v") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
    legend.position="bottom"
  ) +
  scale_fill_brewer("Feature", palette = "Paired") +
  xlab("Samples") +
  ylab("Count")
#ggsave("output/other.count.pdf", width = 16, height = 9)
ggsave("output/other.count.png", width = 16, height = 9)

```

# Heatmap e PCA

```{r}
library(pheatmap)
library(DESeq2)
library(ggrepel)
library(RColorBrewer)

mat <- miRNA %>%
  mutate(across(sample, fct_inseq)) %>%
  arrange(sample, name) %>%
  pivot_wider(names_from = sample, values_from = readCount, values_fill = 0) %>%
  column_to_rownames("name") %>%
  as.matrix

dds <- mat %>%
  DESeqDataSetFromMatrix(colData = coldata, design = ~ 1) %>%
  estimateSizeFactors

dds %>%
  counts %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  write_excel_csv2("output/miRNA.raw.csv")

dds %>%
  counts(normalized = TRUE) %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  write_excel_csv2("output/miRNA.normalized.csv")

pcaData <- dds %>%
  varianceStabilizingTransformation %>%
  plotPCA(intgroup = c("Group", "Treatment", "Day", "Hemolysis"), returnData = TRUE)

percentVar <- attr(pcaData, "percentVar") %>%
  imap_chr(~ paste0("PC", .y, ": ", round(100 * .x), "% variance"))

pcaLimits <- c(-11, 11)

pcaData %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Group, color = Group), size = 5) +
  stat_ellipse(aes(color = Group)) +
  geom_text_repel(aes(label = name), size = 4) +
  scale_shape_manual(values = 16:15) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(percentVar[1], limits = pcaLimits) +
  scale_y_continuous(percentVar[2], limits = pcaLimits) +
  theme_bw()
ggsave("output/PCA.Group.png", width = 16, height = 9)

pcaData %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Group, color = Hemolysis), size = 5) +
  geom_text_repel(aes(label = name), size = 4) +
  scale_shape_manual(values = 16:15) +
  scale_color_brewer(palette = "Reds", na.value = "grey50") +
  scale_x_continuous(percentVar[1], limits = pcaLimits) +
  scale_y_continuous(percentVar[2], limits = pcaLimits) +
  theme_bw()
ggsave("output/PCA.Group.Hemolysis.png", width = 16, height = 9)

pcaData %>%
  inner_join(coldata %>% rownames_to_column("name")) %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Group, color = `RNA Concentration`), size = 5) +
  geom_text_repel(aes(label = name), size = 4) +
  scale_shape_manual(values = 16:15) +
  scale_color_viridis_c(option = "H", direction = -1) +
  scale_x_continuous(percentVar[1], limits = pcaLimits) +
  scale_y_continuous(percentVar[2], limits = pcaLimits) +
  theme_bw()
ggsave("output/PCA.Group.RNA_Concentration.png", width = 16, height = 9)

pcaData %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Group, color = Treatment), size = 5) +
  stat_ellipse(aes(color = Treatment)) +
  geom_text_repel(aes(label = name), size = 4) +
  scale_shape_manual(values = 16:15) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(percentVar[1], limits = pcaLimits) +
  scale_y_continuous(percentVar[2], limits = pcaLimits) +
  theme_bw()
ggsave("output/PCA.Group.Treatment.png", width = 16, height = 9)

pcaData %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Group, color = Day), size = 5) +
  stat_ellipse(aes(color = Day)) +
  geom_text_repel(aes(label = name), size = 4) +
  scale_shape_manual(values = 16:15) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  scale_x_continuous(percentVar[1], limits = pcaLimits) +
  scale_y_continuous(percentVar[2], limits = pcaLimits) +
  theme_bw()
ggsave("output/PCA.Group.Day.png", width = 16, height = 9)

coldata %>%
  ggplot() +
  geom_boxplot(aes(x = Day, y = `RNA Concentration`, fill = Day), show.legend = F) +
  scale_fill_brewer(palette = "Set1", direction = -1) +
  theme_bw()
ggsave("output/boxplot.Day.RNA_Concentration.png", width = 7, height = 7)

mat <- dds %>%
  counts(normalized = TRUE) %>%
  { log2(. + 1) }

mat %>%
  cor %>%
  `^`(2) %>%
  pheatmap(
    annotation_col = coldata %>%
      select(`RNA Concentration`, Hemolysis, Day, Treatment, Group),
    annotation_colors = list(
      `RNA Concentration` = brewer.pal(3, "Spectral"),
      Hemolysis = brewer.pal(5, "Reds") %>% set_names(levels(coldata$Hemolysis)),
      Day = brewer.pal(3, "Set1") %>% set_names(levels(coldata$Day)) %>% rev,
      Treatment = brewer.pal(5, "Set1") %>% set_names(levels(coldata$Treatment)),
      Group = brewer.pal(3, "Set1")[-3] %>% set_names(levels(coldata$Group)) %>% rev
    ),
    color = colorRampPalette(brewer.pal(9, "Blues"))(255),
    main = "Correlation between samples (Pearson R^2)",
    filename = paste0("output/distances.png"),
    width = 16,
    height = 10
  )

hc <- mat %>%
  cor %>%
  { 1 - . } %>%
  as.dist %>%
  hclust

mat %>%
  pheatmap(
    annotation_col = coldata %>%
      select(`RNA Concentration`, Hemolysis, Day, Treatment, Group),
    annotation_colors = list(
      `RNA Concentration` = brewer.pal(3, "Spectral"),
      Hemolysis = brewer.pal(5, "Reds") %>% set_names(levels(coldata$Hemolysis)),
      Day = brewer.pal(3, "Set1") %>% set_names(levels(coldata$Day)) %>% rev,
      Treatment = brewer.pal(5, "Set1") %>% set_names(levels(coldata$Treatment)),
      Group = brewer.pal(3, "Set1")[-3] %>% set_names(levels(coldata$Group)) %>% rev
    ),
    scale = "row",
    cluster_cols = hc,
    filename = "output/heatmap.png",
    width = 20,
    height = 100
  )

mat %>%
  pheatmap(
    annotation_col = coldata %>%
      select(`RNA Concentration`, Hemolysis, Day, Treatment, Group),
    annotation_colors = list(
      `RNA Concentration` = brewer.pal(3, "Spectral"),
      Hemolysis = brewer.pal(5, "Reds") %>% set_names(levels(coldata$Hemolysis)),
      Day = brewer.pal(3, "Set1") %>% set_names(levels(coldata$Day)) %>% rev,
      Treatment = brewer.pal(5, "Set1") %>% set_names(levels(coldata$Treatment)),
      Group = brewer.pal(3, "Set1")[-3] %>% set_names(levels(coldata$Group)) %>% rev
    ),
    #scale = "row",
    cluster_cols = hc,
    filename = "output/heatmap.no_scaling.png",
    width = 20,
    height = 100
  )
  
```

