---
title: "Differential expression of miRNAs"
author: "Marco Bruttini"
date: "19 september 2023"
output: 
  BiocStyle::html_document:
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Intro

All miRNA expression and sample information data are located here:  

`/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-09-18 miRNA count matrix/output`

Longitudinal analysis may be impaired by a strong batch processing effect. See
_2023-09-15 miRNA profiling.pptx_ report.

# Analysis

A minimal low counts filtering was applied:

- miRNAs with at least **5 raw counts (UMI deduped)** in any sample were kept.

## Import data

```{r import}
setwd("/mnt/FdmComune/Progetti LAB/Beta-cell ablation in mice/3) ANALISI/MARCO/2023-09-19 DE miRNAs")

library(tidyverse)

cts <- "../2023-09-05 miRNA count matrix/output/miRNA_wide.csv" %>%
  read_csv2() %>%
  column_to_rownames("name") %>%
  as.matrix

# Experimental design
coldata <- "../_data/Samples.experiment.csv" %>%
  read_csv2(col_types = "fffffcn", locale = locale(decimal_mark = ",")) %>%
  mutate(
    Hemolysis = Hemolysis %>%
      factor(levels = c("No", "Little", "Moderate", "Significant", "Strong")),
    Day = Day %>%
      fct_relabel(~ paste("Day", .x %>% str_pad(2, "left"))),
    Condition = Group %>%
      fct_recode(BCA="β-cell ablation", IR="Insulin resistance") %>%
      paste(Treatment) %>%
      factor,
    rowname = Sample
  ) %>%
  group_by(Condition) %>%
  group_modify(~ .x %>% mutate(Subject = Mouse %>% fct_drop %>% as.integer %>% factor)) %>%
  arrange(Sample) %>%
  column_to_rownames()

```

## Differential expression

### Normalization

```{r "de norm"}
library(DESeq2)
library(ggrepel)
library(ggupset)
library(pheatmap)
library(EnhancedVolcano)
library(BiocParallel)
library(GGally)
library(viridis)
library(ggpointdensity)
library(VennDiagram)
library(ashr)
library(RColorBrewer)

register(SnowParam(12))

cts %>% dim
cts %>% colSums
cts %>% colSums %>% barplot

# Low-counts filtering
cts <- cts[apply(cts, 1, max) >= 10, ]

cts %>% dim
cts %>% colSums
cts %>% colSums %>% barplot

cts %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  write_excel_csv2("output/counts.raw.csv")

dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData = coldata,
  design = ~ 0 + Condition * Day + Subject
) %>%
  DESeq

plotDispEsts(dds)

rld <- dds %>% rlog(blind = FALSE)

rld %>% plotPCA(intgroup = "Group")
rld %>% plotPCA(intgroup = "Treatment")
rld %>% plotPCA(intgroup = "Day")

rld %>%
  assay %>%
  cor %>%
  `^`(2) %>%
  pheatmap(
    annotation_col = coldata %>% select(Day:`RNA Concentration`),
    show_rownames = F,
    show_colnames = F,
    main = "Replicates correlation"
  )

rld <- rld %>%
  assay() %>%
  as.data.frame %>%
  rownames_to_column("miRNA") %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata) %>%
  group_by(Group, Treatment, Day, miRNA)

pointdensity <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_abline(color = "red") +
    geom_pointdensity(...) +
    scale_color_viridis("Density") +
    theme_bw() 
}

rld %>%
  group_by(Group, miRNA) %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(id_cols = miRNA, names_from = Group, values_from = Reads) %>%
  select(!miRNA) %>%
  ggpairs(lower = list(continuous = pointdensity))
ggsave("output/scatter.mean.Group.rlog.png", width = 16, height = 9)

rld %>%
  group_by(Condition, miRNA) %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(id_cols = miRNA, names_from = Condition, values_from = Reads) %>%
  select(!miRNA) %>%
  ggpairs(lower = list(continuous = pointdensity))
ggsave("output/scatter.mean.Condition.rlog.png", width = 16, height = 9)

rld %>%
  group_by(Day, miRNA) %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(id_cols = miRNA, names_from = Day, values_from = Reads) %>%
  select(!miRNA) %>%
  ggpairs(lower = list(continuous = pointdensity))
ggsave("output/scatter.mean.Day.rlog.png", width = 16, height = 9)

rld %>%
  group_by(Mouse, miRNA) %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(id_cols = miRNA, names_from = Mouse, values_from = Reads) %>%
  select(!miRNA) %>%
  ggpairs(lower = list(continuous = pointdensity)) %>%
  ggsave(
    filename = "output/scatter.mean.Mouse.rlog.png",
    width = 32,
    height = 18
  )

rld %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(names_from = Treatment, values_from = Reads) %>%
  pivot_longer(!Group:Control, names_to = "Treatment", values_to = "Treated") %>%
  ggplot() +
  geom_pointdensity(aes(x = Control, y = Treated)) +
  facet_grid(Day ~ Treatment) +
  ggtitle("rlog normalized counts") +
  scale_color_viridis("Density") +
  theme_bw()
ggsave("output/scatter.control_treated.rlog.png", width = 16, height = 9)

rld %>%
  summarise(Reads = Reads %>% mean) %>%
  pivot_wider(names_from = Day, values_from = Reads) %>%
  pivot_longer(!Group:`Day  0`, names_to = "Day", values_to = "Following days") %>%
  ggplot() +
  geom_pointdensity(aes(x = `Day  0`, y = `Following days`)) +
  facet_grid(Day ~ Group + Treatment) +
  ggtitle("rlog normalized counts") +
  xlab("Baseline") +
  scale_color_viridis("Density") +
  theme_bw()
ggsave("output/scatter.baseline_otherdays.rlog.png", width = 16, height = 9)

normalized <- dds %>%
  counts(normalized = TRUE) %>%
  as.data.frame %>%
  rownames_to_column("miRNA")

normalized %>%
  mutate(across(!miRNA, round)) %>%
  write_excel_csv2("output/counts.normalized.csv")

norm.means <- normalized %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Counts") %>%
  inner_join(coldata, by = "Sample") %>%
  group_by(Group, Treatment, Day, miRNA) %>%
  summarise(
    Mean = Counts %>% mean,
    SD = Counts %>% sd
  )

norm.means %>%
  pivot_wider(names_from = Group:Day, values_from = Mean:SD) %>%
  mutate(across(!miRNA, round)) %>%
  write_excel_csv2("output/counts.normalized.group.mean.csv")

norm.means %>%
  slice_max(Mean, n = 100) %>%
  arrange(desc(Mean)) %>%
  mutate(
    miRNA = miRNA %>% fct_inorder
  ) %>%
  ggplot(aes(x = miRNA, y = Mean)) +
  geom_col() +
  geom_errorbar(aes(ymin = pmax(0, Mean - SD), ymax = Mean + SD)) +
  facet_grid(Group + Treatment + Day ~ .) +
  labs(title = "Top 100 expressed miRNAs by group", y = "Mean of normalized counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
ggsave(paste0("output/bar.counts.normalized.top-100.png"), width = 18, height = 32)

norm.means %>%
  ggplot(aes(x = Treatment, y = log2(Mean + 1))) +
  geom_line(aes(group = miRNA), alpha = .1) +
  facet_grid(Day ~ Group, scales = "free_x")
ggsave(paste0("output/line.counts.normalized.group.mean.png"), width = 18, height = 32)

```

### Assessing model performance

We first try a model with interactions because we assume that samples of the
various conditions may respond differently in time:

- `design(dds)`

We want to know if we need to account also for subject variability, by comparing
the full model with a reduced model without Subject variable.

```{r model_1}

design(dds) <- ~ 0 + Day * Condition + Subject

dds.LRT <- dds %>%
  DESeq(test = "LRT", reduced = as.formula(~ 0 + Day * Condition))

res.LRT <- dds.LRT %>%
  results(alpha = 0.05)

sum(res.LRT$padj < 0.05, na.rm = T)

res.LRT[which(res.LRT$padj < 0.05),]

res.LRT %>%
  EnhancedVolcano(
    lab = rownames(.),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    FCcutoff = 0,
    title = "LRT of reduced model without Subject",
    subtitle = bquote(italic("Full = Day * Condition + Subject")),
    caption = paste(
      length(which(.$padj < 0.05)),
      "out of",
      nrow(.),
      "features would better fit the full model"
    )
  )

normalized %>%
  filter(miRNA %in% rownames(res.LRT[which(res.LRT$padj < 0.05),])) %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata) %>%
  ggplot(aes(x = Day, y = log2(Reads + 1), color = Subject)) +
  geom_line(aes(group = Mouse)) +
  geom_point() +
  facet_grid(miRNA ~ Condition, scales = "free") +
  scale_color_brewer("Replicate", palette = "Set1", breaks = 1:4, labels = LETTERS[1:4]) +
  theme_bw()

```

Since only 6 miRNAs would better fit a model with Subject variable, we are
dropping it and continuing with the reduced model.


Now we test for interactions between Condition and Day.

```{r model_2}

design(dds) <- ~ 0 + Day * Condition

dds.LRT <- dds %>%
  DESeq(test = "LRT", reduced = as.formula(~ 0 + Day + Condition))

res.LRT <- dds.LRT %>%
  results(alpha = 0.05)

sum(res.LRT$padj < 0.05, na.rm = T)

res.LRT[which(res.LRT$padj < 0.05),]

res.LRT %>%
  EnhancedVolcano(
    lab = rownames(.),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    FCcutoff = 0,
    title = "LRT of reduced model without interaction",
    subtitle = bquote(italic("Full = Day * Condition")),
    caption = paste(
      length(which(.$padj < 0.05)),
      "out of",
      nrow(.),
      "features would better fit the full model"
    )
  )

normalized %>%
  filter(miRNA %in% rownames(res.LRT[which(res.LRT$padj < 0.05),])) %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata) %>%
  ggplot(aes(x = Treatment, y = log2(Reads + 1))) +
  geom_boxplot(aes(fill = Day)) +
  facet_grid(miRNA ~ Group + Day, scales = "free") +
  scale_fill_brewer("Timepoint", palette = "Reds") +
  theme_bw()

normalized %>%
  filter(miRNA %in% rownames(res.LRT[which(res.LRT$padj < 0.05),])) %>%
  pivot_longer(!miRNA, names_to = "Sample", values_to = "Reads") %>%
  inner_join(coldata) %>%
  ggplot(aes(x = Day, y = log2(Reads + 1), color = Subject)) +
  geom_line(aes(group = Mouse)) +
  geom_point() +
  facet_grid(miRNA ~ Condition, scales = "free") +
  scale_color_brewer("Replicate", palette = "Set1", breaks = 1:4, labels = LETTERS[1:4]) +
  theme_bw()

```

Since only 2 miRNAs would better fit a model with interactions, we are
dropping them and continuing with the simple 
additive model.

Now we test for Day variable.


```{r model_3}

design(dds) <- ~ 0 + Day + Condition

dds.LRT <- dds %>%
  DESeq(test = "LRT", reduced = as.formula(~ 0 + Condition))

res.LRT <- dds.LRT %>%
  results(alpha = 0.05)

sum(res.LRT$padj < 0.05, na.rm = T)

res.LRT[which(res.LRT$padj < 0.05),]

res.LRT %>%
  EnhancedVolcano(
    lab = rownames(.),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    FCcutoff = 0,
    title = "LRT of reduced model without Day",
    subtitle = bquote(italic("Full = Day + Condition")),
    caption = paste(
      length(which(.$padj < 0.05)),
      "out of",
      nrow(.),
      "features would better fit the full model"
    )
  )

```

Since 226 out of 587 miRNAs fit better a full model accounting for
timepoint/batch effect than a reduced model with Condition only, we're keeping
the variable in the design formula.

### Setup DE testing functions

```{r "de setup"}

DE.test <- function(md, dds, ...) {
  print(md$plotTitle)
  print(md$fileName)
  dds %>%
    results(contrast = c("Condition", md$contrast1, md$contrast2), alpha = 0.05) %>%
    lfcShrink(dds = dds, res = ., type = "ashr") %>%
    {
      plotMA(.)
      summary(.)
      .
    } %>%
    as.data.frame %>%
    rownames_to_column("miRNA") %>%
    arrange(padj, pvalue) %>%
    write_excel_csv2(paste0("output/DE/DE.", md$fileName, ".csv")) %>%
    mutate(
      group = md$group,
      test = md$fileName %>% str_remove("^[^.]*\\."),
      direction = if_else(log2FoldChange > 0, "UP", "DOWN"),
      .before = miRNA
    ) %>%
    {
      EnhancedVolcano(
        toptable = .,
        lab = .$miRNA,
        x = 'log2FoldChange',
        y = 'padj',
        pCutoff = 0.05,
        FCcutoff = 0,
        title = md$plotTitle,
        subtitle = bquote(italic("BH adjusted p-values")),
        caption = paste(
          length(which(.$padj < 0.05)),
          "differentially expressed features out of",
          nrow(.),
          "total features"
        )
      )
      ggsave(
        paste0("output/DE/DE.", md$fileName, ".volcano.png"),
        width = 16,
        height = 9
      )
      .
    } %>%
    filter(padj < 0.05) %>%
    {
      if (nrow(.)) {
        select(., miRNA, direction) %>%
          inner_join(normalized, by = "miRNA") %>%
          pivot_longer(!miRNA:direction, names_to = "Sample", values_to = "Reads") %>%
          inner_join(coldata, by = "Sample") %>%
          filter(
            Condition %>% make.names %in% c(md$contrast1, md$contrast2)
          ) %>%
          ggplot(aes(x = Condition, y = log2(Reads + 1))) +
          stat_summary(aes(group = miRNA), fun = mean, geom = "line", linetype = "dashed", alpha = .5) +
          geom_boxplot(aes(fill = Day), position = position_dodge2(padding = 0.2)) +
          facet_wrap(~ direction + miRNA, scales = "free_y") +
          ggtitle(md$plotTitle) +
          #scale_fill_manual(values = brewer.pal(6, "RdYlBu")[c(5, 1)]) +
          scale_fill_manual(values = brewer.pal(6, "RdYlBu")[3:1]) +
          scale_x_discrete(expand = expansion(add = 0.4)) +
          theme_bw()
        ggsave(
          paste0("output/DE/DE.", md$fileName, ".boxplot.png"),
          width = 16,
          height = 9
        )
      }
      .
    }
}

```

### Test: Treatments vs Control accounting for Timepoint

```{r "de test"}

design(dds) <- ~ 0 + Condition + Day

de <- tibble()
    
dds <- dds %>%
  DESeq

resultsNames(dds)

de <- tribble(
  ~group, ~contrast1,     ~contrast2,    ~fileName, ~plotTitle,
  "BCA",  "BCA.DT.005ng", "BCA.Control", "BCA.DT005-vs-CTR", "BCA DT (5ng) vs Control",
  "BCA",  "BCA.DT.015ng", "BCA.Control", "BCA.DT015-vs-CTR", "BCA DT (15ng) vs Control",
  "BCA",  "BCA.DT.120ng", "BCA.Control", "BCA.DT120-vs-CTR", "BCA DT (120ng) vs Control",
  "IR",   "IR.S961",      "IR.Control",  "IR.S961-vs-CTR",   "IR S961 vs Control",
) %>%
  rowwise %>%
  group_map(DE.test, dds = dds) %>%
  bind_rows(de)

```


